Pigeons.jl: Distributed sampling from intractable distributions

### 2024-05-28 paper/PT.tex
Releasing token
Draft of multi thread + distributed combo

\subsection{Distributed implementation}
A distributed implementation of PT is shown in \cref{alg:distributed_PT}
and we describe the details of the implementation below.
For simplicity, we present the algorithm from the perspective that the number of machines available 
is equal to the number of \texttt{Replica}s and distributions, $N$. 
However, Pigeons.jl also allows for the more general case where the number of 
machines is not necessarily equal to $N$.

@@

%In our pseudocode in \cref{alg:distributed_PT} we define the MPI-style 
%functions \texttt{send()}, \texttt{receive!()}, 
%and \texttt{waitall()}. 
%The function \texttt{send()} has three arguments in our pseudocode: the first argument 
%is the object to be sent, the second is the machine number to which the object should 
%be sent, and finally the third argument is a unique tag for this specific send/receive 
%request at this time step and on this machine, generated by \texttt{tag()}.
%The \texttt{tag()} function can be any function that allows one to uniquely identify 
%a tag for a given send/receive request on a given machine at any given time step $t$.
%The function \texttt{receive!()} has the same last two arguments as \texttt{send()} 
%except that the first argument 
%specifies the object to which data should be written directly. Finally, 
%\texttt{waitall(requests)} waits for all initialized MPI \texttt{requests} for a given pair 
%of machines to complete.

@@

    \For{$t$ {\bf in} 1, 2, \dots, $T$}
		  \If{$t$ is even} \Comment{choose between even or odd swap}
		    \State $P \gets \{i: 1 \le i < N, i \text{ is even} \}$
		  \Else
		    \State $P \gets \{i: 1 \le i < N, i \text{ is odd} \}$
		  \EndIf

      \State $x_t \gets \texttt{local\_exploration}(\pi_\texttt{chain}, x_{t-1})$
      \State $c_t \gets \texttt{chain}$ 
      \If{$\texttt{chain} \in P$}
        \State $\texttt{partner} \gets \texttt{chain}+1$
      \ElsIf{$\texttt{chain}-1 \in P$}
        \State $\texttt{partner} \gets \texttt{chain}-1$
      \Else 
      	\State $\texttt{partner} \gets \texttt{chain}$
      \EndIf
      \State $\texttt{to\_machine} \gets  \texttt{permuted\_get}(\texttt{dist\_array}, \texttt{partner})$
      \State $\texttt{my\_stat} \gets \texttt{swap\_stat}(x_t, $\texttt{chain}, $\texttt{partner} )$  
      \State $\texttt{partner\_stat} \gets \texttt{transmit}(\texttt{my\_stat}, \texttt{to\_machine})$
      \If{$\text{swap\_accepted}(\texttt{my\_stat}, \texttt{partner\_stat})$}
        \State $\texttt{chain} \gets \texttt{partner\_chain}$
      \EndIf
      \State $\texttt{permuted\_set!}(\texttt{dist\_array}, \texttt{chain}, j)$ 
		\EndFor
    \State \Return $\{(x_t, c_t)\}_{t=1}^T$
	\end{algorithmic}
  \caption{Distributed PT on machine $j$ (one replica per machine)}
  \label{alg:distributed_PT}
\end{algorithm}

>>>>

\subsection{Distributed implementation}
A distributed implementation of PT is shown in \cref{alg:distributed_PT}
and we describe the details of the implementation below.
For pedagogy, we first present the algorithm in the special case where the number of machines available 
is equal to the number of \texttt{Replica}s and distributions, $N$. We outline 
how Pigeons removes this restriction in \cref{sec:relax}.
However, Pigeons.jl also allows for the more general case where the number of 
machines is not necessarily equal to $N$.

@@

@@

    \For{$t$ {\bf in} 1, 2, \dots, $T$}

     \State $x_t \gets \texttt{local\_exploration}(\pi_\texttt{chain}, x_{t-1})$
    \State $c_t \gets \texttt{chain}$

		  \If{$t$ is even} \Comment{choose between even or odd swap}
		    \State $\texttt{direction} \gets +1$
		  \Else
		    \State $\texttt{direction} \gets -1$
		  \EndIf
		  \State $\texttt{partner} \gets \texttt{chain} + \texttt{direction}$

      \If{$\texttt{partner} = 0$} 
        \State $\texttt{partner} \gets 1$ \Comment{Encode boundaries as self-interaction}
      \ElsIf{$\texttt{partner} = N + 1$}
        \State $\texttt{partner} \gets N$
      \EndIf
	    \State $\texttt{my\_stat} \gets \texttt{swap\_stat}(x_t, $\texttt{chain}, $\texttt{partner} )$  
      \State $\texttt{to\_machine} \gets  \texttt{permuted\_get}(\texttt{dist\_array}, \texttt{partner})$
      \State $\texttt{partner\_stat} \gets \texttt{transmit}(\texttt{my\_stat}, \texttt{to\_machine})$
      \If{$\text{swap\_accepted}(\texttt{my\_stat}, \texttt{partner\_stat})$}
        \State $\texttt{chain} \gets \texttt{partner\_chain}$
      \EndIf
      \State $\texttt{permuted\_set!}(\texttt{dist\_array}, \texttt{chain}, j)$ 
		\EndFor
    \State \Return $\{(x_t, c_t)\}_{t=1}^T$
	\end{algorithmic}
  \caption{Distributed PT on machine $j$ (one replica per machine)}
  \label{alg:distributed_PT}
\end{algorithm}

\subsubsection{Load balancing}
\label{sec:relax}

So far we have assumed, for pedagogy, that the number of 
machines $M$ is equal to the number of chains $N$.
This restriction is not present in Pigeons, so we now describe how 
the code base generalizes \cref{alg:distributed_PT}. 
First, the $N$ replicas are shared between the $M$ machines so that 
each machine contains either $\lceil N/M \rceil$ or $\lfloor N/M \rfloor$ 
replicas. Each machine's replicas are called the local replicas. 
The allocation of replicas to machine is fixed over the whole execution of the 
algorithm and maintained using the \texttt{LoadBalance} struct. 
Second, the code performing local exploration is enclosed in a 
loop over the local replicas, and if the options ``\texttt{multithreaded = true}'' 
is used in the \texttt{pigeons(...)} call, the exploration loop is performed in a 
\texttt{@threads for} loop. 
Finally, the functions used in the communication, 
\texttt{permuted\_get}, \texttt{permuted\_set} \texttt{transmit}, etc., 
are vectorized to accommodate that \texttt{chain} and 
\texttt{partner} are no longer integers but rather sets of integers. 
In particular, the permuted distributed array is designed so that each 
machine is responsible for tracking machine allocations for 
$\lceil N/M \rceil$ or $\lfloor N/M \rfloor$ chains instead of a single one.
